Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2022-09-30 paper/paper-infer-supp.tex
save need discu
integrate Clara's comments
continue changes

\begin{figure}[p]
	\centerline{\includegraphics[width=\textwidth]{ukbb_h2_p}}
	\caption{Estimates from LDpred2-auto for either the SNP heritability $h^2$ or the polygenicity $p$ for all 248 phenotypes defined from the UK Biobank. The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept. \label{fig:ukbb_h2_p}}
\end{figure}

@@

\begin{figure}[p]
	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
	\caption{LDpred2-auto estimates for UKBB phenotypes using either the HapMap3 or HapMap3+ sets of variants. Only 154 phenotypes with more than 25 chains kept when using the HapMap3 variants are represented here. Red dashed lines represent the 1:1 line. The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept. The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap. \label{fig:hm3_plus}}
\end{figure}

@@

@@

\begin{figure}[p]
\centerline{\includegraphics[width=\textwidth]{median_postp}}
\caption{Per-variant median probabilities of being causal across 169 UKBB phenotypes. The HapMap3+ set of variants is used. Only phenotypes with more than 25 chains kept are used here. \label{median_postp}}
\end{figure}

>>>>

\begin{figure}[p]
	\centerline{\includegraphics[width=\textwidth]{ukbb_h2_p}}
	\caption{Estimates from LDpred2-auto for either the SNP heritability $h^2$ or the polygenicity $p$ for all 248 phenotypes defined from the UK Biobank. Estimates of $h^2$ are constrained to be at least 0.001, and at least $10^{-5}$ for $p$. The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept. \label{fig:ukbb_h2_p}}
\end{figure}

@@

@@

\begin{figure}[p]
\centerline{\includegraphics[width=\textwidth]{median_postp}}
\caption{Per-variant median probabilities of being causal across 169 UKBB phenotypes. Variants with a median probability larger than 0.01 were mapped to genes using R package rsnps (only the largest per peak). The HapMap3+ set of variants is used. Only phenotypes with more than 25 chains kept are used here. \label{median_postp}}
\end{figure}

@@

### 2022-09-30 paper/paper-infer.tex
continue changes

Most traits and diseases are heritable. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
Understanding how heritable and polygenic a phenotype is can provide valuable clinical insight and inform future research efforts.
Therefore, many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome. 
These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$ used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$ that can inform about negative selection, \cite{schoech2019quantification}), SuSiE (per-variant $p$ used for fine-mapping, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).

@@

First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$), and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
The $\alpha$ estimate can become unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
Then, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
We recall that calibrated per-variant probabilities of being causal can be used for fine-mapping purposes  \cite[]{wang2020simple}.
Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).

We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a threshold corresponding to the prevalence.
Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$). The main difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2012better}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.

@@

Then, to investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).

We then investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde40\% more variants on top of HapMap3 variants recommended to use for LDpred2 (Methods). As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
In particular, a much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.

@@

In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
We compute per-block (local) $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by a single block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
The outlier block explaining a much larger heritability contains the HLA region.
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; interestingly, these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a negative correlation of -5.5\% with the probabilities of being causal.
To verify that this does not correspond to regions of low LD, we compute LD scores; the probabilities above have a correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.

@@

Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 40\% larger to offer a better coverage of the genome.
This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
Using the new HapMap3+ set also improves predictive performance by an average of 6.1\% here, and particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
This means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.

Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable. However, estimates of $h^2$ and $r^2$ seem always reliable, except for maybe height for which they are probably overestimated.
We think this is likely due to assortative mating \cite[]{border2022assortative,yengo2022saturated}. 
Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics; future work could focus on correcting this issue. [TODO: OTHERS?]

>>>>

Most traits and diseases are heritable. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
Understanding how heritable and polygenic a phenotype is can provide valuable clinical insight and inform future research efforts.
A third parameter, often called $\alpha$ or $S$, can also be used to quantify the effect of negative selection on complex traits \cite[]{schoech2019quantification,zeng2021widespread}.
Therefore, many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$. 
These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$ used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$, \cite{schoech2019quantification}), SuSiE (per-variant $p$ used for fine-mapping, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).

@@

First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$), and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
The $\alpha$ estimate can become unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
We recall that calibrated per-variant probabilities of being causal can be used for fine-mapping purposes  \cite[]{wang2020simple}.
Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).

We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a threshold corresponding to the prevalence.
Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$). The main difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2011estimating}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.

@@

Then, to investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes [TODO: CHECK IF ARE THE SAME] for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).

We then investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde40\% more variants on top of HapMap3 variants recommended to use for LDpred2, to improve the coverage of this set (Methods). As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
In particular, a much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
	\caption{LDpred2-auto estimates for UKBB phenotypes using either the HapMap3 or HapMap3+ sets of variants. Only 154 phenotypes with more than 25 chains kept when using the HapMap3 variants are represented here. Red dashed lines represent the 1:1 line. The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept. The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap. \label{fig:hm3_plus}}
\end{figure}

@@

In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
We compute local per-block $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by a single block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
The outlier block explaining a much larger heritability contains the HLA region.
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; interestingly, these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a negative correlation of -5.5\% with the probabilities of being causal.
To verify that this does not correspond to regions of low LD, we compute LD scores; the median probabilities of being causal have a correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.

@@

Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 40\% larger to offer a better coverage of the genome.
This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
Using the new HapMap3+ set also improves predictive performance by an average of 6.1\% here, and particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
This means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more localized set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.

Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable. However, estimates of $h^2$ and $r^2$ seem always reliable, except for maybe height for which they are probably overestimated.
We think this is likely due to assortative mating \cite[]{border2022assortative,yengo2022saturated}. 
Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics; future work could focus on correcting this issue.
